---
title: "Assignment_2_4726_schetry2"
author: "Saurav Prem Kaushik Chetry"
date: "9/26/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Preparation  
### Load required R packages and apply proper transformations on the Boston Housing Data.
```{r}
mypackages = c("MASS", "glmnet")   # required packages
tmp = setdiff(mypackages, rownames(installed.packages()))  # packages need to be installed
if (length(tmp) > 0) install.packages(tmp)
lapply(mypackages, require, character.only = TRUE)

```

```{r}
set.seed(4726)
myData = Boston
names(myData)[14] = "Y"
iLog = c(1, 3, 5, 6, 8, 9, 10, 14);
myData[, iLog] = log(myData[, iLog]);
myData[, 2] = myData[, 2] / 10;
myData[, 7] = myData[, 7]^2.5 / 10^4
myData[, 11] = exp(0.4 * myData[, 11]) / 1000;
myData[, 12] = myData[, 12] / 100;
myData[, 13] = sqrt(myData[, 13]);
```


```{r}

# Move the last column of myData, the response Y, to the 1st column.
X = as.matrix(myData[, -14])
y = myData$Y
lam.seq = c(0.30, 0.2, 0.1, 0.05, 0.02, 0.005)
```

## Some Technical Details  
### We use the following function to solve the Lasso estimate for βj given other coefficients fixed; see the derivation in Coding2.pdf.

```{r}

one_var_lasso = function(r, x, lam){
    xx = sum(x^2)
    xr = sum(r * x)
    b = (abs(xr) - lam/2)/xx
    b = sign(xr) * ifelse(b > 0, b, 0)
    return(b)
    #print(b)
}
```

## CD for Lasso  
### Implement the Coordinate Descent algorithm for Lasso. Some part of the function MyLasso is blocked here, but your submission should include all code used to produce your results.

```{r}

MyLasso = function(X, y, lam.seq, maxit = 50) {
    
    # X: n-by-p design matrix without the intercept 
    # y: n-by-1 response vector 
    # lam.seq: sequence of lambda values 
    # maxit: number of updates for each lambda 
    # Center/Scale X
    # Center y
  
    n = length(y)
    p = dim(X)[2]
    nlam = length(lam.seq)
  
    ##############################
    # YOUR CODE: 
    # Record the corresponding means and scales
    # For example, 
    # y.mean = mean(y)
    # yc = centered y
    # Xs = centered and scaled X
    ##############################
    
    y.mean = mean(y)
    yc = y - y.mean
    # Xs = scale(X, center = TRUE, scale = TRUE)
    Xs = matrix(0,nrow = nrow(X), ncol = ncol(X))
    for (i in 1:ncol(X)){

      Xs[,i] = (X[,i]- mean(X[,i]))/sqrt(sum((X[,i]- mean(X[,i]))^2)/(n))

    }


    
    # Initilize coef vector b and residual vector r
    b = rep(0, p)
    r = yc
    B = matrix(nrow = nlam, ncol = p + 1)
    
    # Triple nested loop
    for (m in 1:nlam) {
        lam = 2 * n * lam.seq[m]
        for (step in 1:maxit) {
            for (j in 1:p) {
                r = r + (Xs[, j] * b[j])
                b[j] = one_var_lasso(r, Xs[, j], lam)
                r = r - Xs[, j] * b[j]
            }
          
        }
        #print(b)
        B[m, ] = c(0, b)
        
    }
    #print((B))
    ##############################
    # YOUR CODE:
    # Scale back the coefficients;
    # Update the intercepts stored in B[, 1]
    ##############################
    
    for (m in 1:nlam){
      
      total = 0
      
      for (j in 1:p){
        #print(m)
        #print(j+1)
        #print(X)
        total = total + B[m,j+1]*mean(X[,j])/sqrt(sum((X[,j]- mean(X[,j]))^2)/(n))
        B[m,j+1] = B[m,j+1]/sqrt(sum((X[,j]- mean(X[,j]))^2)/(n))
        
      }
      B[m,1] = y.mean - total
    }
    #print(t(B))
    return(t(B))
}
```

```{r}
myout = MyLasso(X, y, lam.seq, maxit = 50)
rownames(myout) = c("(Intercept)", colnames(X))
myout


```

```{r}
#lam.seq = c(0.30, 0.2, 0.1, 0.05, 0.02, 0.005)
lasso.fit = glmnet(X, y, alpha = 1, lambda = lam.seq)
coef(lasso.fit)
```
### Compare the accuracy of my algorithm against the output from glmnet. The maximum difference between the two coefficient matrices is less than 0.005.
```{r}
max(abs(coef(lasso.fit) - myout))
```


