---
title: "Assignment_3_4765_schetry2"
author: "Saurav Prem Kaushik Chetry"
date: "9/17/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
mypackages = c("splines", "glmnet")   # required packages
tmp = setdiff(mypackages, rownames(installed.packages()))  # packages need to be installed
if (length(tmp) > 0) install.packages(tmp)
lapply(mypackages, require, character.only = TRUE)
```

## Prepare your function  

### Write your own function to use LOO-CV and GCV to select the optimal span for loess. If you do not know where to start, you can follow the structure below to prepare your functions.

```{r}
lo.lev <- function(x1, sp){
  # x1: feature vector of length n
  # sp: a value for "span"
  
  n = length(x1);
  lev = rep(0, n)
  
  ##############################################
  # YOUR CODE: Compute the diagonal entries of the 
  #            smoother matrix S and 
  #            store it in a vector "lev"
  # Tip: check how we compute the smoother matrix
  #      for smoothing spline models
  ##############################################

  S = matrix(0, n, n)
  
  for (i in 1:n) {
    y = rep(0, n)
    y[i] = 1
    S[, i] = loess(y ~ x1, span = sp, control = loess.control(surface = "direct"))$fitted
  }
  lev = diag(S)
  
  return(lev)
}

#custom function for loocv
cv.lo = function(x1,y1,y1_hat,sp){
  
  diagS = lo.lev(x1, sp)
  n = length(y1)
  cv = rep(0,n)
  
  for(i in 1:n){
    
    cv[i] = ((y1[i] - y1_hat[i])/(1 - diagS[i]))^2
    
  }
  return(mean(cv))
}

# custom function for gcv
gcv.lo = function(x1,y1,y1_hat,sp){
  diagS = lo.lev(x1, sp)
  n = length(y1)
  gcv = rep(0,n)
  
  for(i in 1:n){
    
    gcv[i] = ((y1[i] - y1_hat[i])/(1 - mean(diagS[i])))^2
    
  }
  return(mean(gcv))
  
}

onestep_CV <- function(x1, y1, sp){
  
  ##############################################
  #  YOUR CODE: 
  #  1) Fit a loess model y1 ~ x1 with span = sp, and extract 
  #     the corresponding residual vector
  #  2) Call lo.lev to obtain the diagonal entries of S
  #  3) Compute LOO-CV and GCV using formula from lecture notes
  #    [lec_W5_NonlinearRegression.pdf] page 33. 
  ##############################################
  
  mod.lo = loess(y1 ~ x1, span = sp, control = loess.control(surface = "direct"))
  cv = cv.lo(x1,y1,mod.lo$fitted,sp)
  gcv = gcv.lo(x1,y1,mod.lo$fitted,sp)
  
  return(list(cv = cv, gcv = gcv))
}

myCV <- function(x1, y1, span){
  # x1: feature vector of length n
  # y1: response vector of length n
  # span: a sequence of values for "span"
  
  m = length(span)
  cv = rep(0, m)
  gcv = rep(0, m)
  
  for(i in 1:m){
    val = onestep_CV(x1, y1, span[i])
    cv[i] = val$cv
    gcv[i] = val$gcv
  }
  return(list(cv = cv, gcv = gcv))
}
```

## Test your function  

### Test your function with data "Coding3_Data.csv."

```{r}
mydata = read.csv(file = "Coding3_Data.csv")
#head(mydata)
dim(mydata)
```

```{r}
plot(mydata$x, mydata$y, xlab="", ylab="")
```

### Create a grid of values for span: 15 values that are equally spaced between 0.20 and 0.90. Call your function myCV to compute the corresponding LOO-CV and GCV.

```{r}
span1 = seq(from = 0.2, by = 0.05, length = 15 )
cv.out = myCV(mydata$x, mydata$y, span1)
```

## Print out your results  

### Print your results on LOO-CV and GCV. Both achieve their minimal at 0.5.

```{r}
myout = data.frame(CV = cv.out$cv, 
                   GCV = cv.out$gcv, 
                   span = span1)
myout
```

```{r}
myout$span[myout$GCV == min(myout$GCV)]
```

```{r}
myout$span[myout$CV == min(myout$CV)]
```


## Plot the fitted curve   

### Plot the data (red circles), the true curve (gray) and the fitted curve (blue dashed line) using the optimal span.

```{r}
spangcv.min = 0.5
plot(mydata$x, mydata$y, xlab="x", ylab="y", col="red");
fx = 1:50/50;
fy = sin(12*(fx+0.2))/(fx+0.2)
lines(fx, fy, col=8, lwd=2)

f = loess(y ~ x, mydata, span = spangcv.min)
lines(fx, predict(f, data.frame(x = fx), surface = "direct"), 
      lty=2, lwd=2, col="blue")

legend("topright", legend = c("data", "true", "fitted"), lty = c(NA,1,2), pch = c(1,NA,NA), col = c("red", "gray", "blue"), cex = 0.7)
```



